\iffalse
One-page printable interview handout for Akgül et al. 2026
arxiv:2605.06241 — "Rethinking RL for LLM Reasoning"
Written per the write-latex skill's rules:
  - no fontspec
  - \section* (unnumbered) by default
  - pdflatex-clean
  - comments only inside \iffalse ... \fi
\fi

\documentclass[10pt]{article}

\usepackage[letterpaper, margin=0.6in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{titlesec}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  citecolor=blue!50!black,
  urlcolor=blue!50!black
}

\titleformat{\section}{\large\bfseries}{}{0pt}{}
\titlespacing*{\section}{0pt}{6pt}{2pt}

\setlist[itemize]{leftmargin=*, topsep=1pt, itemsep=1pt}

\pagenumbering{gobble}

\begin{document}

\begin{center}
  {\Large\textbf{Akg\"ul et al.\ 2026 --- Rethinking RL for LLM Reasoning}}\\[2pt]
  {\small Sparse Policy Selection, Not Capability Learning \quad\textbar\quad arXiv:2605.06241 \quad\textbar\quad \href{https://github.com/pleyva2004/rethinking-rl-llm-reasoning-sparse-policy-selection}{my study repo}}
\end{center}

\section*{One-Liner}
RL post-training for math reasoning reranks $\sim$2\% of tokens at high-entropy decision points toward alternatives the base model already had in its top-5; an offline contrastive fine-tune gated on base entropy matches full RL at $\sim 1000\times$ less compute.

\section*{What's Novel}
\begin{itemize}
  \item Token-level mechanistic claim with a causal probe: oracle correction at reranked positions recovers RL's pass@1 exactly; random controls do not.
  \item Teacher-free locator: base entropy alone ($\tau \approx 1.4$ nats) identifies the intervention sites without ever consulting the RL model.
  \item Constructive alternative: \emph{ReasonMaxxer} drops the online RL loop, keeps GRPO's per-group advantage normalisation, applies CE only at decision points, KL-anchors elsewhere.
  \item RL's parametric delta is rank-32 LoRA on QKVO (0.27--0.49\% of params); rank-8 on $W_O$ alone nearly suffices.
\end{itemize}

\section*{Math in 60 Seconds}
Token-level predictive entropy under the base model
\[
  H_t = -\sum_{v} \pi_{\text{base}}(v \mid q, o_{<t}) \log \pi_{\text{base}}(v \mid q, o_{<t}).
\]
Decision-point set: $D = \{t : H_t > \tau\}$. Per-rollout advantage (from GRPO):
\[
  A_i = \frac{r_i - \bar r}{\sigma_r + \epsilon}, \quad r_i \in \{0,1\}.
\]
ReasonMaxxer loss:
\[
  \mathcal{L} = -\sum_i A_i \!\!\sum_{t \in D^{(i)}}\!\! \log \pi_\theta\big(o_t^{(i)} \,\big|\, q, o_{<t}^{(i)}\big) \;+\; \lambda \!\!\sum_{t \notin D^{(i)}}\!\! \mathrm{KL}\big( \pi_{\text{base}} \,\big\|\, \pi_\theta \big).
\]
Empirical: reranked fraction 1--4\%; the RL-promoted token sits at mean rank $\approx 2.2$ under the base; entropy at reranked positions is $5$--$12\times$ that at unchanged ones.

\section*{What I'd Push Back On}
\begin{itemize}
  \item ``Within base top-5'' is measured on already RL-trained checkpoints; partial circularity until verified on a held-out, truly untouched base.
  \item Threshold $\tau$ is tuned per setup. ``No RL needed'' partially trades the optimisation loop for a scalar grid-search.
  \item All wins are on math with verifiable scalar rewards. Open-ended tasks, agentic tool use, or tasks where the right intervention is \emph{low}-entropy fall outside the licensed scope --- and the framework predicts its own failure there.
  \item ``1000$\times$ cheaper'' compares to full RL training only; ReasonMaxxer still needs labeled-correctness rollouts. Accounting could be sharper.
\end{itemize}

\section*{My One Proposed Extension}
A Lipschitz bound on rank displacement. The paper observes the RL-promoted token sits at mean rank $\approx 2.2$ across all four base/RL pairs --- a striking regularity left unexplained. I conjecture that under bounded GRPO learning rate $\eta$ and per-step KL-to-reference budget $\beta$, the expected rank displacement satisfies $\mathbb{E}[\Delta_t \mid t \in R] \leq C\eta/\beta$. The derivation routes through the policy-ratio bound under the standard KL-clipped GRPO update \citep{shao2024grpo} and a Lipschitz argument on the softmax-rank-maximiser. Falsifiable with a single training-run sweep across $\eta$ at fixed $\beta$, checking linear scaling. Closes the loop between the empirical mean-rank observation and the underlying optimisation dynamics, and would extend cleanly to predict where the top-5 assumption breaks under higher-learning-rate regimes.

\emph{Full proposal set: \texttt{05-improvements.tex} in the study repo.}

\section*{My One Question}
\emph{If the entire RL behavioural delta is captured by a rank-32 LoRA on $W_O$, why train at all? What's the obstacle to replacing the training stage with a search over inference-time steering vectors applied only at high-entropy positions?}

\end{document}
